# Basic Data Manipulation with tidyverse {#sec-manipulation_examples}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width="0.5\\linewidth", fig.align="center")
library( tidyverse )

```

Let's use the following simple table and run through various tidyverse commands. This is not a complete reference! See the text of the textbook for further details.

```{r}
library( tidyverse )
table1
```

## `filter()` (Grab the rows you want, 5.1)

```{r}
filter( table1, year > 1999 )
```

Remember, if you want to save the results of your command, you need to put it in a new variable, like so:

```{r}
my.table <- filter( table1, year > 1999 )
my.table
```

When you do something like this, you should see it appear in your workplace.

## `arrange()` (Sort your rows the way you want, 5.2)

```{r}
arrange( table1, cases )
arrange( table1, desc( country ), population )
```

## `select()` (Grab the columns you want, 5.3)

```{r}
select( table1, country, population )
```

## `mutate()` (Make new variables out of your old ones, 5.4)

```{r}
table1 <- mutate( table1, case.per.1000 = 1000 * cases / population )
table1
```

(We will use this new variable later, so I am saving it in our table)

## `group_by()` and `summarize` (Summarize your data by subgroup, 5.6)

```{r}
tbl <- group_by( table1, country )
summarize( tbl, av.pop = mean( population ), av.cases = mean( cases ) )
```

Same thing, with the pipe!

```{r}
table1 %>% group_by( country ) %>%
    summarize( av.pop = mean( population ), av.cases = mean( cases ) )
```

## Special: grouped mutates (Making new variables within subgroups, 5.6)

This combo is for doing things like group mean centering your data:

```{r}
table1 %>% group_by( year ) %>% mutate( case.per.1000.cent = case.per.1000 - mean( case.per.1000 ) )
```

And a plot

```{r}
ggplot( table1, aes( x=year, y=case.per.1000, col=country ) ) + geom_line() + geom_point()
```
