---
title: "Using mice to handle categorical data"
author: "Miratrix"
editor_options: 
  chunk_output_type: console
---

```{r,include=FALSE}
library(tidyverse)
library( mice )
knitr::opts_chunk$set(echo = TRUE,
          fig.width = 5,
          fig.height = 3,
          out.width = "75%",
           message = FALSE,
          fig.align = "center")
options(list(dplyr.summarise.inform = FALSE))
theme_set( theme_classic() )
```

Non-continuous data is a bit tricky because we can't directly do mean imputation, since there is nothing to take a mean of.
That said, all the ideas shown in the prior intro to missing data chapter can be extended to non-continuous (categorical) variables.

We have a few options:

 - Add a Missing category
 - Do mean imputation on dummy variables
 - Use mice to impute values.  This is probably the easiest option.

We walk through these next, using the `nhanes` data (another built-in dataset in the mice package) for our illustration.
In `nhanes`, one of the variables is an `age` variable that is coded as an ordered categorical variable (e.g. 20-39, 40-59, etc).
Another is the binary `hyp` variable.
To illustrate imputation, we are just going to delete some of the `age` data so we can try to impute it back.

```{r}
## load data and create some missing values for an ordered categorical
## variable

  data(nhanes2)
  nhanes2$age[1:5] = NA
  head(nhanes2) 
```

Ok, we have our data.
Next we go through the way of handling it.

## Adding a missing category

Adding a missing category is a common approach for categorical variables, and it is easy to do.

For example, for the `age` variable, we can simply add a new level called `"Missing"` to
cover any `NA` observations, using `fct_explicit_na()` from the `forcats`
package (loaded automatically with `tidyverse`):

```{r}
  nhanes2_cat <- nhanes2 %>%
    mutate(age = fct_na_value_to_level(age, level="Missing"))
  
  levels(nhanes2_cat$age)
  table(nhanes2_cat$age)
```

The `age` variable now has a "`Missing`" level, so it can
be passed directly into a regression or other model.
The same approach works for the binary `hyp` variable:
```{r}
  nhanes2_cat <- nhanes2_cat %>%
    mutate(hyp = fct_na_value_to_level(hyp, level="Missing") )
  
  table(nhanes2_cat$hyp)
```

The appeal of this approach is its simplicity: no imputation model is required,
and the `"Missing"` category can sometimes capture meaningful signal (e.g., if a
certain type of respondent systematically skips a question).
The downside is that it adds a category whose interpretation is not always clear,
and if the data are not MCAR it could introduce bias in downstream analysis just like listwise deletion
would.


## Mean imputation on dummy variables

A common approach is to convert a categorical variable into a set of dummy (0/1)
variables and then do ordinary mean imputation on each dummy.

This works especially naturally for a binary variable like `hyp`, which takes
values `"yes"` and `"no"`.
Converting `hyp` to a 0/1 dummy and then mean-imputing is equivalent to replacing
each missing value with the sample proportion of `"yes"` responses among
non-missing cases:
```{r}
  nhanes2_dum <- nhanes2 %>%
    mutate(hyp_num = as.integer(hyp == "yes"))
  
  head(nhanes2_dum)
  
  hyp_mean <- mean(nhanes2_dum$hyp_num, na.rm = TRUE)
  hyp_mean
  
  nhanes2_dum <- nhanes2_dum %>%
    mutate(hyp_imp = if_else(is.na(hyp_num), hyp_mean, as.double(hyp_num)))
  
  head(nhanes2_dum)
  summary(nhanes2_dum$hyp_imp)
```

The originally-missing `hyp` rows now have the value `r round(hyp_mean, 3)`,
which is the proportion of hypertensive respondents in the observed data.

For an ordered categorical variable like `age`, you would create a separate dummy
for each level (dropping one to avoid perfect collinearity) and mean-impute each
one.

However, just as with continuous variables, mean imputation has well-known limitations: it attenuates relationships
between variables and underestimates variance.
For categorical variables, it also produces fractional imputed values (e.g., 0.4) that do not correspond to any valid category.
E.g., what does it mean to say our first observation has 0.23 for "having hypertension"?


## Using Mice to impute values

`Mice` will automatically detect and handle non-continuous (e.g. dichotomous or categorical) variables on its own.
That said, it will also allow you to specify the type of variable you are imputing yourself, if you want more control over the process.

Imputing missing values is easy:
```{r}
  imp.cat <- mice(nhanes2, m = 1, print=FALSE)     
  full.cat = mice::complete(imp.cat)        
  head(full.cat)
```

We now have categorical values in our `age` column and `hyp` column, along with the continuous imputations for the other variables.

The `mice` package uses a different model for each variable, and the model can depend on what kind of data we are imputing.
We can check which imputation method `mice` used as so:

```{r}
  imp.cat$method
```

`mice` used the `polyreg` imputation method for the variable *age*, which means it treated it as an unordered categorical variable.
We also see that a logistic regression was used for our binary variable `hyp`.
But we know our *age* variable is an ordered variable: higher values categories signified older age.
We might want to use a model that takes that ordering into account.

One very important method often used is "polr" (Proportional odds model) which
handles _ordered_ categories, instead of unordered.  This allows it to say some
general range of categories is more or less likely, given the other variables.

In general, the more structure you can impose (based on your domain knowledge)
means better imputation.

More broadly, to know what methods for imputation are available, see the various methods described under mice's help file (or check out the long but readable mice paper by @van2011mice (also see the Canvas site for the reference).


We can manually force `mice` to treat *age* as an ordered as so:

```{r}
  imp.cat2 <- mice(nhanes2, 
                   meth=c("polr","pmm","logreg","pmm"),
                   m=1, print=FALSE)
  head(mice::complete(imp.cat2, 1))
  imp.cat2$method
```


### A coding trick to get the defaults

As a coding tip, we can use mice with `maxit=0` to get what the defaults are first, and
then we can easily change the ones we want, and run mice for real.
This is useful if we have a lot of variables, and do not want to type in all the defaults:

```{r}
imp <- mice(nhanes2, maxit = 0 )
imp$method
mth = imp$method
mth[1] = "polr"
mth
```

And now run as before:
```{r}
imp <- mice( nhanes2, method=mth, print=FALSE )
```

Easy as pie!



## References




