---
title: "Prediction and Plotting"
author: "Miratrix"
date: "2/7/2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width="0.5\\linewidth", fig.align="center")
library( tidyverse )

```

# Overview

To make nice plots from funky, curved models, the easiest way is to use the model to predict values for different covariate combinations, store all of this in a dataframe, and then use `ggplot` to plot your results using `geom_line` which will give you nice smoothed curves through the points.

This document hilites some useful methods for doing this. See R for DS 23.3. We need the tidyverse and the `modelr` package:

```{r, message=FALSE, warnings=FALSE}
library( tidyverse )
library( modelr )
```

## Some important functions

Some important functions are: `add_predictions()` and `add_residuals()`

Also see `gather_pred()` to compare models.

Here is an example using the iris dataset

```{r}
head(iris)
ggplot( iris, aes( x = Petal.Length, y = Petal.Width, col=Species ) ) +
    geom_point()
```

Let's fit a simple linear model ignoring species

```{r}
M0 = lm( Petal.Length ~ Petal.Width, data=iris )
summary( M0 )
```

Let's fit a linear model with a species interaction:

```{r}
M1 = lm( Petal.Length ~ Petal.Width * Species, data=iris )
summary( M1 )
```

Now let's add some predictions to our original data:

```{r}
my.iris = iris
my.iris = add_predictions(iris, M1 )
head( my.iris )
```

## More on the model matrix

See `model_matrix`, a slightly better version of `model.matrix` in that it gives you a tibble rather than a matrix, so it is easier to deal with.

```{r}
model_matrix( Petal.Length ~ Petal.Width * Species, data=iris )
```

## `data_grid()`

This function, in the `modelr` package, given a dataframe, will generate a simple dataframe with one row for each possible combination of variables you want:

```{r}
library( modelr )
my.df = data.frame( A=c(1,2,1,2), B=c(10,20,20,20), C=c(10,20,30,40) )
data_grid( my.df, A, B )
```

You can also get "typical values" (usually the median or most commonly occuring) for things not on your grid using a modeling argument:

```{r}
mod <- lm(Petal.Length ~ Petal.Width * Species, data=iris)
data_grid(iris, .model = mod)
data_grid(iris, Petal.Width = seq_range(Petal.Width, 9), .model = mod)
```

Similar is `seq_range()` that gives you a nice sequence of numbers over a range of values

```{r}
X = rnorm( 10000 )
seq_range( X, 5 )

# drop 5% of data on tails
seq_range( X, 5, trim=0.05 )

# use pretty numbers
seq_range( X, 5, trim=0.05, pretty=TRUE )
```

You can combine with `data_grid` to get a nice data frame of points to use for your "canonical points" for prediction or whatnot:

```{r}
my.df = data.frame( A=rnorm(1000), B=rexp(1000), C=rnorm(1000,mean=5) )
data_grid( my.df, 
           A=seq_range(A,n=3),
           B=seq_range(B,n=3) )
```

# Plotting

The above is useful for plotting:

```{r}
grid = data_grid(iris, Petal.Width = seq_range(Petal.Width, 9), .model = mod)
grid2 = add_predictions(grid, M1, "M1" )
grid2 = add_predictions(grid2, M0, "M0" )
head( grid2 )

# more simply done with this
grid2 = spread_predictions( grid, M0, M1 )
head( grid2 )

# but for plotting, we want long form
grid2 = gather_predictions( grid, M0, M1 )
head( grid2 )
```

Ok, now let's plot:

```{r}
ggplot( grid2, aes(x= Petal.Width, y= pred, col=Species, lty=model ) ) +
    geom_line(lwd=2)
```

Note that our plot is not perfect: the three lines for the three M0 species predictions are all the same and thus we see only blue (the green and red get blotted out). We could get fancy:

```{r}
grid2 = spread_predictions( grid, M0, M1 )
head( grid2 )
ggplot( grid2, aes(x= Petal.Width ) ) +
    geom_line( aes( y=M0 ), col="black", lwd=2 ) +
    geom_line( aes( y=M1, col=Species ), lwd=2, lty=2 )
```

## Bonus: Interactions and curved lines

The above generalizes. Let's add a quadratic model:

```{r}
M2 = lm( Petal.Length ~ (Petal.Width + I(Petal.Width^2)) * Species, data=iris )
grid2 = add_predictions( grid, M2 )
ggplot( grid2, aes(x= Petal.Width, y= pred, col=Species ) ) +
    geom_line( lwd=1)

```
