---
title: "Prediction and Visualization with Models"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      out.width="0.5\\linewidth", 
                      fig.align="center",
                      warning = FALSE,
                      message = FALSE)
library( tidyverse )
library(modelr)
theme_set(theme_bw())
```


To make nice plots of the predictions from models that have complex curves and relationships, the easiest way is to use the model to predict values for different covariate combinations, store all of this in a dataframe, and then use `ggplot` to plot your results using `geom_line` which will give you nice smoothed curves through the points. This document highlights some useful methods for doing this. We need the tidyverse and the `modelr` package

Prediction, and making plots from predictions, are foundational skills for data analysis, especially as models become more complex and interactions or non-linear terms are introduced.

We will use the `iris` dataset, as it has some strong relationships that make the ideas behind these tools more salient. Here is the raw data:

```{r}
ggplot(iris, aes(x=Petal.Length, y=Petal.Width, col=Species)) +
  geom_point()
```

To visualize we go through a series of steps:

1) Fit a model to describe the relationship between variables
2) Generate a grid of values to predict over
3) Add predictions to the grid
4) Make the vizualization, plotting our predictions vs. our grid values.

Related reading: R for Data Science (1st Edition), Section 23.3 *(Sections 23.1 and 23.2 are optional but helpful)*

## Fit a model

We first fit a a model to describe how petal length can be predicted by petal width and species. Our model allows the length-width relationship to be different for the different species.

```{r}
M1 <- lm(Petal.Length ~ Petal.Width * Species, data=iris)
summary( M1 )
```

## Generate a Prediction Grid

We often want to understand what our model is saying by predicting outcomes over combinations of input variables. We do this by first making a new dataset of the combinations of inputs we want to explore, and then we use our fit model to predict for all these combinations.

`data_grid()`, in the `modelr` package, is the tool for this.
It will generate a simple dataframe with one row for each possible combination of variables you want.  We will then predict for each of these rows.

Here is a toy example: see how duplicates are removed:

```{r}
library( modelr )
my.df = data.frame( A=c(1,2,1,2), B=c(10,20,20,20), C=c(10,20,30,40) )
data_grid( my.df, A, B )
```

For our actual data we have:
```{r}
grid = iris %>%
  data_grid( Petal.Width, Species )
grid
```

You can also get "typical values" (usually the median or most commonly occurring) for things not on your grid using a modeling argument:

```{r}
mod <- lm(Petal.Length ~ Petal.Width * Species, data=iris)
data_grid(iris, .model = mod)
data_grid(iris, Petal.Width = seq_range(Petal.Width, 9), .model = mod)
```

A tool that is also good is `seq_range()` that gives you a nice sequence of numbers over a range of values

```{r}
X = rnorm( 10000 )
seq_range( X, 5 )

# drop 5% of data on tails
seq_range( X, 5, trim=0.05 )

# use pretty numbers
seq_range( X, 5, trim=0.05, pretty=TRUE )
```

You can combine with `data_grid` to get a nice data frame of points to use for your "canonical points" for prediction or whatnot:

```{r}
my.df = data.frame( A=rnorm(1000), B=rexp(1000), C=rnorm(1000,mean=5) )
data_grid( my.df, 
           A=seq_range(A,n=3),
           B=seq_range(B,n=3) )
```



## Add Predictions to the Grid

This is easy, like so:

```{r}
grid <- grid %>% 
  add_predictions(M1)
head(grid)
```

## Plot Predicted Trends

We can then plot our predictions for each of our species:

```{r}
ggplot(grid, aes(x=Petal.Width, y=pred, 
                 col=Species, group=Species)) +
  geom_line()
```

## Extensions

### Residuals and Diagnostics

We can add `residuals` to our data for model diagnostics:

```{r}
iris <- add_residuals(iris, M1)
ggplot(iris, aes(x=Petal.Width, y=resid, col=Species)) +
  geom_jitter() +
  geom_ref_line(h=0, col="orange", size=0.5)
```

### Multiple Models

Let’s try multiple models and compare their predictions. We will have a simple model with no interactions, and a more complex one that allows curved relationships within each species.

```{r}
M0 <- lm( Petal.Length ~ Petal.Width + Species, data=iris )
M2 <- lm( Petal.Length ~ Species *(Petal.Width + I(Petal.Width^2)), data=iris )

grid = iris %>%
  data_grid( Petal.Width, Species ) %>%
  gather_predictions( M1, M0, M2 )
grid
```

And now plot:

```{r}
ggplot(grid, aes(x=Petal.Width, y=pred, col=Species ) ) +
  facet_wrap(~ model) +
  geom_line()
```

### Fancy Plotting with spread_predictions

The `spread_predictions()` function can be used to make each model a separate column:

```{r}
grid <- iris %>%
  data_grid( Petal.Width, Species )
grid_spread = spread_predictions(grid, M0, M1)
grid_spread

ggplot(grid_spread, aes(x=Petal.Width)) +
  facet_wrap( ~ Species ) +
  geom_line(aes(y=M0), col="black", lwd=1) +
  geom_line(aes(y=M1, col=Species), lwd=1, lty=1)
```

## Handy Functions Recap

| Function | Use |
|--------------------------------------------|:---------------------------|
| `data_grid()` | Create a grid of input values for prediction |
| `add_predictions()` | Add predicted values from a model to a dataset |
| `add_residuals()` | Add model residuals to original data |
| `gather_predictions()` | Combine predictions from multiple models for plotting |
| `spread_predictions()` | Spread predictions into separate columns |
| `seq_range()` | Create evenly spaced values over a variables range |

## Final Thoughts

These techniques make model interpretation and communication much easier. Being able to **visualize predictions and model trends** can bring clarity, especially when dealing with interactions or nonlinearities.

Practice using these tools and adapt them to your own datasets!
